```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# The Data Science Process

As a data scientist, what steps are involved in answering a question with data? In this course, we aim to frame everything in the context of the data science process.

In the following chapters and courses we will cover each part of this process more in-depth.

### The Parts of a Data Science Project

Data science process generally follow these steps:

The first part of the a data science process is **forming a question**. Sometimes you'll have a dataset in mind that you will explore to form this question, and other times, you will have a question that you will find some data to help you answer.

- **Form a question** - What is the question you hope to answer with data?
- **Get the data** - To answer this question you'll need data!
- **Clean the data** - Datasets are almost never ready to analyze from the get go. As a data scientist, we'll need to do some cleaning and set up steps to get the data where we need it before we can really dig into it.
- **Explore and plot the data** - After your data is clean, you'll want to do some initial exploration. What do these data look like? Plots are a great way to visual your data and start to get an idea of how the data might relate to your initial question.
- **Get stats** - You can use statistics and models to try to use your data to get to the bottom of your question.
- **Share results** - Given everything you are seeing with your data, how do we interpret this? What's our conclusion when it comes to our initial question?

```{r, fig.alt = "Data Science process involves", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/114QYFmKuJ2M5E3tlBw8Gwu2xI09tb8gnrKrNkMY9CG4/edit#slide=id.p")
```

In the upcoming chapters and courses we will further dive into each of these steps!

### A Data Science Project Example

For this example, we’re going to use an example analysis from a machine learning engineer named [Hazma](https://medium.com/@Hamza.b86). The specific project we’ll be working through here is from 2018 and titled [Exploratory Data Analysis in R of Global Data from GapMinder](https://towardsdatascience.com/exploratory-data-analysis-in-r-of-global-data-from-gapminder-13c5de5bee70). To get the most out of this lesson, click on that link and read through Hazma’s post. Once you’re done, come on back to this lesson and read through the breakdown of this post.

#### Form a question

When setting out on a data science project, it’s always great to have your question well-defined. Additional questions may pop up as you do the analysis, but knowing what you want to answer with your analysis is a really important first step. Hazma’s question is included in bold in his post. Highlighting this makes it clear that he’s interested in answer the following question:

> Is there a correlation between GDP per Capita and prevalence of HIV in the 15–49 age bracket? And if yes, how strong is that correlation?

#### Get the data

To answer this question, Hazma collected data from Gapminder, a free data source from the World Bank. This dataset included measures from many countries of Gross Domestic Product (GDP) per capita, as well as the prevalence of HIV in those nations. GDP is an economic indicator signifying a country’s collective wealth.

#### Data Analysis

As explained in the blog post, Hazma expected to discover a negative correlation between GDP per capita and HIV prevalence. In other words, he theorized that less wealthy nations experience higher rates of HIV relative to their wealthier peers. Completing this analysis by hand, this would be a time-intensive and inefficient. Thankfully, by writing code in R, Hazma was able to answer his question with a few simple data transformations.
It’s not important at this point in time to fully understand the techniques Hazma used in his analysis, but it is important to know that after getting the data together, the next step is figuring out how to reshape the data in order to answer your question. To answer his question, Hazma needed to merge the two data variables—GDP per capita and HIV prevalence—into a single dataframe.

#### Clean the data

To prepare the data for analysis, Hazma used R to transform the data from Gapminder into a format that could answer his question. He describes this process in his blog post so that others could see what he did and repeat his steps if they wanted.

#### Explore and plot the data

Data science projects often involve writing code and generating figures that aren’t always included in your final results. This data exploration is part of the data science process too. Familiarizing yourself with the characteristics of the dataset can help build confidence in your results, and it can inspire new lines of inquiry you might not have considered otherwise. For example:

> How many countries are included in this dataset?
> Are there any periods of time when certain countries have missing data? > Are particular regions more susceptible to incomplete data history?
> What sort of bias might missing data introduce to the analysis?

This line of questioning might not have a direct impact on the answer to your original question. However, Data Scientists believe that this kind of curiosity (and sometimes skepticism) is not a distraction from our core work, but rather, a vital component of a robust analysis.

#### Get stats

After data exploration, Hazma began to further analyze the data. The first thing he did was look at GDP per capita for 275 countries. The data reveals a rising trend in Global GDP values from 1960 to 2011.

![Biggest Drop Table](https://docs.google.com/presentation/d/1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8/export/png?id=1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8&pageid=g39818b4900_0_36)

In looking at the results of this analysis, the first five years appeared peculiar to Hilary Parker. (It's always good to consider whether or not the results were what you were expecting, from any analysis!) None of them seemed to be names that were popular for long periods of time. To see if this hunch was true, Hilary plotted the percent of babies born each year with each of the names from this table. What she found was that, among these "poisoned" names (names that experienced a big drop from one year to the next in popularity), all of the names other than Hilary became popular all of a sudden and then dropped off in popularity. Hilary Parker was able to figure out why most of these other names became popular, so definitely read that section of her post! The name, Hilary, however, was different. It was popular for a while and then completely dropped off in popularity.


![14 most poisoned names over time](https://docs.google.com/presentation/d/1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8/export/png?id=1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8&pageid=g39818b4900_0_47)

To figure out what was specifically going on with the name Hilary, she removed names that became popular for short periods of time before dropping off, and only looked at names that were in the top 1000 for more than 20 years. The results from this analysis definitively show that Hilary had the quickest fall from popularity in 1992 of any female baby name between 1880 and 2011. ("Marian"'s decline was gradual over many years.)


![39 most poisoned names over time, controlling for fads](https://docs.google.com/presentation/d/1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8/export/png?id=1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8&pageid=g39818b4900_0_53)

#### Interpret and communicate results

The final step in this data analysis process was, once Hilary Parker had answered her question on her computer, it was time to share it with the world. An important part of any data science project is effectively communicating the results of the project. Hilary did so by writing a wonderful blog post that communicated the results of her analysis, answered the question she set out to answer, and did so in an entertaining way.

Additionally, it's important to note that most projects build off someone else's work. It's *really* important to give those people credit. Hilary accomplishes this by:

* linking to a [blog post](http://stuartbuck.blogspot.com/2003/09/hillary-is-most-poisoned-baby-name-in.html) where someone had asked a similar question previously
* linking to the [Social Security website](https://www.ssa.gov/OACT/babynames/) where she got the data
* linking to a link about where she [learned about web scraping](http://syntaxi.net/2013/01/20/storyboard/)

### What you can build using R

Hilary's work was carried out using the R programming language. Throughout the courses in this series, you'll learn the basics of programming in R, exploring and analyzing data, and how to build reports and web applications that allow you to effectively communicate your results. To give you an example of the types of things that can be built using the R programming and suite of available tools that use R, below are a few examples of the types of things that have been built using the data science process and the R programming language - the types of things that you'll be able to generate by the end of this series of courses.

#### Prediction Risk of Opioid Overdoses in Providence, RI

Masters students at the University of Pennsylvania set out to predict the risk of opioid overdoses in Providence, Rhode Island. They include [details on the data they used, the steps they took to clean their data, their visualization process, and their final results](https://pennmusa.github.io/MUSA_801.io/project_5/index.html). While the details aren't important now, seeing the process and what types of reports can be generated is important. Additionally, they've created a [Shiny App](https://jordanbutz.shinyapps.io/directory/), which is an interactive web application. This means that you can choose what neighborhood in Providence you want to focus on. All of this was built using R programming.


![Prediction of Opioid Overdoses in Providence, RI](https://docs.google.com/presentation/d/1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8/export/png?id=1SNT3SYuWJhjRYx7VmyFKWkuxESEx5THt-mWJ7Mx5Cr8&pageid=g39818b4900_0_93)

### Other Cool Data Science Projects

The following are smaller projects than the example above, but data science projects nonetheless! In each project, the author had a question they wanted to answer and used data to answer that question. They explored, visualized, and analyzed the data. Then, they wrote blog posts to communicate their findings. Take a look to learn more about the topics listed and to see how others work through the data science project process and communicate their results.

* [Predicting movie ratings with IMDb data and R](http://re-design.dimiter.eu/?p=767), by [Dimiter Toshkov](http://re-design.dimiter.eu/)
* [Where to Live in the US](http://www.masalmon.eu/2017/11/16/wheretoliveus/), by [Maelle Salmon](https://masalmon.eu/)
* [Sexual Health Clinics in Toronto](https://sharlagelfand.netlify.com/posts/tidying-toronto-open-data/), by [Sharla Gelfand](https://sharlagelfand.netlify.com/about/)
