# (PART\*) Sharing Results {-}

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# **Sharing results**

```{r, fig.alt = "Data Science process", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/114QYFmKuJ2M5E3tlBw8Gwu2xI09tb8gnrKrNkMY9CG4/edit#slide=id.g145cde68be4_0_154")
```

## Learning Objectives

Through the completion of this section our goal is that you will be able to:

- Understand what reproducibility means
- Value the importance of making your analyses reproducible
- Understand the importance of version control
- Navigate GitHub comfortably
- Create a GitHub repository and link it to RStudio Cloud
- Use the basic git command to version control an analysis
- Recognize the different methods available for communicating data science results
- Create a data analysis results report including all the important components
- Give a presentation using Google slides and be able to tailor this presentation for a particular audience
- Create a blog that fosters conversation
- Recognize common expectations for participating in meetings
