```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Pushing and Pulling Changes

Now that you have learned how to create a repository and clone it, it's time to start using repositories to manage projects and work with collaborators. In this lesson you will learn about new git commands for publishing changes so that others can see them. You will also learn a git command for incorporating changes made by others in your local copy of a repository.

### Opening a PR

A pull request (sometimes abbreviated PR) is a way of being able to review changes before you incorporate them into your main, more polished product. It is a highly effective system for doing code review and otherwise communicating about your data analysis to increase its reproducibility.

```{r, fig.align='center', echo = FALSE, fig.alt= "A direct changes set up: Doesn’t allow for others to easily review and give feedback. Is more likely to lead to mistakes being published and disrupting course experience. Rationale of the changes made are less likely to be clear. Is quicker in the short-term. A pull request model: Encourages collaboration and feedback. Adds an extra safety net so mistakes are less likely to be published and disrupt course experience. Encourages more clear documentation and tracking about the changes being made. Requires more time and effort to do it right; but this generally has long-term payoffs!"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1MNHf8JpolaEP_vQ_kB-1xRBF9wo3haCArRu117hBoHA/edit#slide=id.g2288a8ec77f_0_88")
```

Pull requests are based on copies of the project repository that are called `branches`.
When we do work on a project we will want to do it somewhere that is separate from our main set of code. Branches are yet another copy of the code that is used for developing purposes.

<div class = "dictionary">
- **pull request** - a method of working on and incorporating file changes in a way that allow things to be reviewed and discussed on GitHub.
- **branch** - another copy and version of your project that you can work from and create a pull request.
</div>

#### Creating a new branch

A `main` branch is where you will keep your best, most vetted version of the project. Your `main` branch will already exist on your project when you create it. When others come to look at your project, they will see the `main` branch first.

Other branches are generally made with the purpose of eventually having them reviewed and polished to the point that they are to add their changes to the `main` branch.

```
pr_init(branch = "new_branch_name")
```

This branch will be called "new_branch_name" but you should call your branch whatever would be representative of the work you will be doing on this. For example, if you are adding documentation, you could call your branch "adding-docs".

This brings us to another point. For keeping yourself organized as well as making it easier for others to follow, it's best to keep a branch and subsequent pull request focused on one task. If one pull request tries to do too many things, it will be more difficult for it to be communicated and reviewed properly.

Now that you have created a branch, you are ready to make changes to your files.

Now make any change to a file for the purposes of this tutorial. For example, you could add a sentence to your README file to explain what this project will be. Then save the file change.

In RStudio, if we go to the `Git` tab we should see that the file we changed has an `M` next to it to signify it has been modified. In order to officially add these changes that we've made to our branch, we need to `commit` them.  

<div class = "dictionary">
- **commit** - the action of officially adding a file change to a branch.
</div>

To do this, we first need to check the box(es) next to the files we'd like to commit then we click the commit button.

```{r, fig.align='center', echo = FALSE, fig.alt= "To commit a file change, first go to the Git tab. Then you should see the file that you change listed. In this case we see that we changed the README.md file and it has a blue M next to it to signify that it has been modified. Next we should check the box for the file changes we’d like to add. Next we click Commit in order to add these changes."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1MNHf8JpolaEP_vQ_kB-1xRBF9wo3haCArRu117hBoHA/edit#slide=id.g2288a8ec77f_0_150")
```

After clicking the commit button, a new window will pop up that shows us the changes we are committing. In this window, we will need to add a commit message. Again, we should try to write something informative about what we were doing to these files we are committing. Then after we've written this commit message, we need to click `Commit` in this window.


```{r, fig.align='center', echo = FALSE, fig.alt= "After you click ‘Commit’ a new window should pop up. Green lines indicate added bits while red indicate things that were deleted. We’ll write an informative commit message about the changes we are adding and then click ‘Commit’ again. "}
ottrpal::include_slide("https://docs.google.com/presentation/d/1MNHf8JpolaEP_vQ_kB-1xRBF9wo3haCArRu117hBoHA/edit#slide=id.g2288a8ec77f_0_164")
```

Now our files are on our branch!

When we'd like these changes to be online on GitHub for others to see, we can `push` these changes. To push changes means to send them online to GitHub.

<div class = "dictionary">
- **push** - the action of sending a branch and its file changes to GitHub so it is online where others can see it.  
</div>

With the `usethis` package, we can use the `pr_push()` function.

```
pr_push()
```

This should open a new window on GitHub that will look like this:

```{r, fig.align='center', echo = FALSE, fig.alt= "On the new window that opens upon using pr_push(), you can click Create pull request to create your pull request. "}
ottrpal::include_slide("https://docs.google.com/presentation/d/1MNHf8JpolaEP_vQ_kB-1xRBF9wo3haCArRu117hBoHA/edit#slide=id.g2288a8ec77f_0_177")
```

It will also print out some messages like this:
```
✔ Pushing local 'branch_name' branch to 'origin' remote.
• Create PR at link given below
✔ Opening URL 'https://github.com/username/repository_name/compare/branch_name'
```

On this window, click create pull request. This will bring you to another page on GitHub where you can create your pull request. On this page you will want to describe the changes you are making with this pull request.

You should include information like:

1) The background behind the changes you are making. What is the problem you are solving? Link to any relevant conversations.
2) What changes are you making specifically and how do these address the problem?
3) What work is left to be done?
4) What help could you use from others? Is there something in particular you would like to be looked at? Is there something not yet added that should be added?

```{r, fig.align='center', echo = FALSE, fig.alt= "On this page, you should leave a description of what changes you are making with this pull request. Try to be as informative and to the point as possible. It’s particularly important to describe the background of the changes you are making as well as what work is left to be done and what you could use other’s help with."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1MNHf8JpolaEP_vQ_kB-1xRBF9wo3haCArRu117hBoHA/edit#slide=id.g2288a8ec77f_0_184")
```

Now if you continue to work with your files, you can go through the same steps of:

1) Making a file change
2) Check the box(es) next to the file changes you'd like to add
3) Commit the file
4) Add a commit message
5) Commit the changes
6) Push the changes by running `pr_push()` (You can also click the green arrow to push if you prefer).


You can repeat these steps as many times as you need until you feel the file changes on your PR page are ready for someone else to review. If you don't have someone else on your team to review the changes, you can alternatively let them sit for a day or two and let "future You" review your file changes. Future you or others on your project might have a fresh take on these files. We will talk more about code review in a later chapter.

Code review is perhaps the most powerful tool for making reproducible analyses. And now that you know how to make pull requests on GitHub, you have an excellent platform and system for version controlling and tracking, and reviewing your files! Congrats!

### Merging a pull request

In the `Files changed` tab, you or others can leave comments about your file changes. This is an excellent way to document the rationale of these file changes as well as discuss any alternatives.

After some back and forth discussion (whether this discussion be between you and yourself or someone else) you may decide that the code on this pull request is ready for primetime! In other words, it is ready to be incorporated into the `main` branch.

Recall that we said the `main` branch is the most polished and readily viewable version of your project. To bring the changes in your pull request into the main branch, we will need to perform an action called a `merge`.

 This step of incorporating changes into a branch and combining two branches together is called `merging`. The goal of a merge is to combine two branches in such a way that keeps your desired changes from both copies.

<div class = "dictionary">
- **main** - the branch name that is typically used for the main, most polished and live version of your project that others will be brought to first.
- **merge** - combining the files from two branches into one.
</div>

On your pull request page, scroll to the bottom to the big green button where it says `Merge pull request`. Be careful to not press this button before you are sure that everyone on your team is ready.

After you've clicked it, hooray! You've completed your first merge of a pull request.

It may feel like we've described a lot of steps, but getting into the GitHub workflow and utilizing its benefits is all about habits! We promise it will increase the reproducibility of your analyses if you are able to stick with this process!

### Practice

To get more practice with these concepts, as we know they are not the easiest, we suggest you go practice at [GitHub Learning Lab](https://github.com/apps/github-learning-lab). Here you can practices your git commands and get feedback as to what you've done correctly and where you steered off course. It's a great way to get better at using git commands before you move onto the quiz!
