```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# About the Authors {-}

These credits are based on our [course contributors table guidelines](https://github.com/jhudsl/OTTR_Template/wiki/How-to-give-credits).

&nbsp;
&nbsp;

|Credits|Names|
|-------|-----|
|**Pedagogy**||
|Content Author(s) | [Shannon Ellis], [Candace Savonen], [Aboozar Hadavand], [Leslie Myint], [Leonardo Collado-Torres]|
|Content Contributor(s)| [Sarah McClymont], [Leah Jager]|
|Content Editor(s)/Reviewer(s) | [Davon Person]|
|Content Director(s) |[Jeff Leek]|
|**Production**||
|Content Publisher(s)| [Candace Savonen]|
|Content Publishing Reviewer(s)| |
|**Technical**||
|Course Publishing Engineer(s)| [Candace Savonen]|
|Template Publishing Engineers|[Candace Savonen], [Carrie Wright]|
|Publishing Maintenance Engineer|[Candace Savonen]|
|Technical Publishing Stylists|[Carrie Wright], [Candace Savonen]|
|Package Developers ([ottrpal])|[John Muschelli], [Candace Savonen], [Carrie Wright]|
|Package Developers ([swirlify])|[Sean Kross]|
|**Art and Design**||
|Illustrator(s)| [Shannon Ellis], [Candace Savonen], [Aboozar Hadavand] |
|**Funding**||
|Funder(s)| Johns Hopkins Bloomberg School of Public Health, Bloomberg Philanthropies |
|Funding Staff| [Ashley K Johnson]|

&nbsp;

```{r, echo = FALSE}
devtools::session_info()
```

<!-- Author information -->
[Ashley K Johnson]: https://www.linkedin.com/in/ashleykgjohnson
[Davon Person]: https://www.linkedin.com/in/davon-person-1ba973194
[Aboozar Hadavand]: https://www.ahadavand.com/
[Shannon Ellis]: https://www.shanellis.com/
[John Muschelli]: https://johnmuschelli.com/
[Candace Savonen]: https://www.cansavvy.com/
[Carrie Wright]: https://carriewright11.github.io/
[Jeff Leek]: https://jtleek.com/
[Leslie Myint]: https://lmyint.github.io/
[Leonardo Collado-Torres]: https://lcolladotor.github.io/
[Sean Kross]: https://seankross.com/
[Sarah McClymont]: https://www.linkedin.com/in/sarahmcclymont/
[Leah Jager]: https://lrjager.github.io/

<!-- Links -->

[ottrpal]: https://github.com/jhudsl/ottrpal
